---
title: "APM 1205: Formative Assessment 1"
subtitle: "Part A - Initial Examination of the Data"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
library(dplyr)
library(ggplot2)
```

## Part A – Initial Examination of the Data

### a) Data Inspection and Measurement Roles

```{r load-data}
# Load the data directly from the online file
url <- "https://raw.githubusercontent.com/resbaz/r-novice-gapminder-files/master/data/gapminder-FiveYearData.csv"
gapminder_all <- read.csv(url)

# Filter for the year 2007
gapminder_2007 <- gapminder_all %>% filter(year == 2007)

# Show rows and columns
dim(gapminder_2007)
```

* **Dimensions:** The dataset contains **142 observations** (countries) and **6 variables**.
* **Measurement Roles:** 
  * `lifeExp` is the **Response Variable ($Y$)**.
  * `gdpPercap` is the **Explanatory Variable ($X$)**.

***

### b) Numerical Summaries

```{r summaries}
# Generate basic summary statistics
gapminder_2007 %>% 
  select(lifeExp, gdpPercap) %>% 
  summary()
```

* **Center and Spread:**
  * **Life Expectancy:** The median is **71.94 years**, which sits higher than the mean of **67.01 years**. It ranges across a spread from **39.61** to **82.61 years**.
  * **GDP per Capita:** The median is **\$6,124**, while the mean is much higher at **\$11,680**. It has a massive spread ranging from **\$277** to **\$49,357**.

***

### c) Graphical Distributions (Histograms)

```{r histograms, fig.show="hold", out.width="50%"}
# Histogram for Life Expectancy
ggplot(gapminder_2007, aes(x = lifeExp)) +
  geom_histogram(fill = "blue", color = "white", bins = 20) +
  labs(title = "Distribution of Life Expectancy (2007)", x = "Years", y = "Count")

# Histogram for GDP per Capita
ggplot(gapminder_2007, aes(x = gdpPercap)) +
  geom_histogram(fill = "orange", color = "white", bins = 20) +
  labs(title = "Distribution of GDP per Capita (2007)", x = "USD", y = "Count")
```

* **Shape & Skewness:**
  * **Life Expectancy:** The distribution is **left-skewed** (negatively skewed), as most countries cluster toward the higher end.
  * **GDP per Capita:** The distribution is heavily **right-skewed** (positively skewed), with most countries packed at low income levels and a few rich nations stretching far to the right.

***

### d) Boxplots and Outliers

```{r boxplot, fig.width=6, fig.height=3}
# Boxplot for GDP per Capita to check for outliers
ggplot(gapminder_2007, aes(x = gdpPercap)) +
  geom_boxplot(fill = "lightgray") +
  labs(title = "Boxplot of GDP per Capita (2007)", x = "USD")
```

* **Observations Deserving Further Examination:** The boxplot for `gdpPercap` shows several individual points (dots) past the right whisker boundary. These are extreme **high-income outliers** that will exert an undue pull on a standard linear regression model.

